\MakeUppercase{Linear versus centred Colouring Numbers}\thanks{This research was partly funded by NSERC.}

### 2021-10-07 lin-vs-cen.tex
Some easy observations that improve exponent to 39

% Document-specific commands and math operators
\DeclareMathOperator{\tw}{tw}
\DeclareMathOperator{\chicen}{\chi_{\mathrm{cen}}}
\DeclareMathOperator{\chilin}{\chi_{\mathrm{lin}}}
\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\vor}{Vor}

@@

The \emph{centred chromatic number} $\chicen(G)$ of $G$ is the minimum integer $c$ such that $G$ has a centred $c$-colouring.  The \emph{linear chromatic number} of $G$ is the minimum integer $\ell$ such that $G$ has a linear $\ell$-colouring.  Since every path in $G$ is a connected subgraph of $G$, every centred colouring of $G$ is also a linear colouring of $G$, so $\chilin(G)\le\chicen(G)$.

@@

These notes are an attempt to improve the bound in \cref{kun_obrien_general} in the general case as well as for some special classes of graphs.

As a first step, we attempt to improve the third part of the argument to show that, if $G$ contains a $k\times k$ grid minor, then $\chilin(G)\in \Omega(k)$.  This, by itself, reduces the exponent in \cref{kun_obrien_general} from $190$ to $95$.  Next, we observe that the first two parts of the argument are both ``heavy hammers'' and that, for some special graph classes, much lighter hammers suffice.  For example, if we consider only planar graphs, then it is well known that any planar graph $G$ of treewidth $k$ contains an $\Omega(k\times k)$ grid minor.  This already reduces the exponent further (for planar graphs) to $5$.  If one could prove a similar result for the first ``heavy hammer'' then this would show that, for any planar graph $G$, $\chicen(G)\in O(\chilin(G))$.

>>>>

% Document-specific commands and math operators
\DeclareMathOperator{\tw}{tw}
\DeclareMathOperator{\td}{td}
\DeclareMathOperator{\chicen}{\chi_{\mathrm{cen}}}
\DeclareMathOperator{\chilin}{\chi_{\mathrm{lin}}}
\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\vor}{Vor}

@@

The \emph{centred chromatic number} $\chicen(G)$ of $G$ is the minimum integer $c$ such that $G$ has a centred $c$-colouring.  We note that an easy argument shows that centered chromatic number and treedepth are equivalent, i.e., $\td(G)=\chicen(G)$ for every graph $G$. The \emph{linear chromatic number} of $G$ is the minimum integer $\ell$ such that $G$ has a linear $\ell$-colouring.  Since every path in $G$ is a connected subgraph of $G$, every centred colouring of $G$ is also a linear colouring of $G$, so $\chilin(G)\le\chicen(G)$.

@@

These notes are an attempt to improve the bound in \cref{kun_obrien_general} in the general case as well as for some special classes of graphs. First we point out that some new results can already reduce the exponent from $190$ to $39$ and eliminate the polylog factor:

A recent result of \citet[Theorem~1.4]{czerwinski.nadara.ea:improved} shows that if $b$ is the maximum treedepth of any tree contained in $G$ then $\td(G)\le b\cdot\tw(G)$ (and the proof is easy).  If we take $\tw(G)=k^{38}$ and $b=k$ then, for any graph $G$ with $\td(G)\ge k^{39}$, one of the following cases applies:
\begin{enumerate}
    \item $\tw(G)\ge k^{38}$: In this case the proof follows Steps~2 and 3 as before to show that $\chilin(G)\in\Omega(k)$.
    \item $G$ contains a tree $T$ of treedepth $k$.  The same paper \cite[Lemma~1.6]{czerwinski.nadara.ea:improved} shows that any tree $T$ of treedepth $k$ contains a subcubic subtree of treedepth $\Omega(k)$.
    A theorem of \citet{kun.obrien.ea:polynomial} shows that, for any subcubic tree $T$, $\chicen(T) \le (\log_2 3)\cdot\chilin(T)$.  Therefore, in this case $\chilin(G)\ge\chilin(T)\ge \chicen(T)/\log_2 3 = \Omega(k)$.
\end{enumerate}

This gives the following improvement to \cref{kun_obrien_general}
\begin{thm}[\citet{kun.obrien.ea:polynomial}]\label{kun_obrien_general2}
  For any graph $G$, $\chicen(G)\in O(\chilin^{39}(G))$.
\end{thm}

As a next step, we would like to improve the third part of the argument to show that, if $G$ contains a $k\times k$ grid minor, then $\chilin(G)\in \Omega(k)$.  This, by itself, reduces the exponent in \cref{kun_obrien_general2} from $39$ to $20$.  We observe that the first two parts of the argument are both ``heavy hammers'' and that, for some special graph classes, much lighter hammers suffice.  For example, if we consider only planar graphs, then it is well known that any planar graph $G$ of treewidth $k$ contains an $\Omega(k\times k)$ grid minor.  This already reduces the exponent in \cref{kun_obrien_general2} further (for planar graphs) to $3$ ($2$ if we can prove the result for grid minors).  If one could prove a similar result for the first ``heavy hammer'' then this would show that, for any planar graph $G$, $\chicen(G)\in O(\chilin(G))$.
